Debugging by \textit{lastChange}

### 2010-11-01 trunk/papers/pldi2011/sigplanconf-template.tex
Claude's comments - 2

A bug is \textit{reproducible} for a developer when the developer can
start from a determined initial state, operate on the program with a
list of actions, and reproduce the symptoms of the bug. The details of
the execution can change each time we re-execute the buggy program,
but the buggy result is the same.  The reproducibility of the bug
means that the defect is very unlikely to depend on the order of
events during the execution. When the developer selects a value and
asks for \textit{lastChange}, they are expressing an hypothesis that
the value is related to the defect. Therefore the reproducibility of
the bug makes it very unlikely that the \textit{lastChange} will
depend upon order of events during execution.

Since \textit{lastChange} does not rely on determinism, some bugs that
confound breakpoint debuggers will be easier to find with
\textit{lastChange}.  Consider the example in 
Figure ~\ref{fig:counter-example}. It shows a function which processes an
array. The array contains numbers except one item which is
\texttt{undefined} and it causes a bug in line 30. There is a call to
\texttt{randomPermutation} function in line 11 which randomly
permutes the array item. So at every execution the
\texttt{undefined} item will be in a new place. Calling
\textit{lastChange} on \texttt{x} at the place bug happens, gives a
point which shows line 14. Although this point exists at every
re-execution but it can not be identified by a conditional breakpoint.

@@

Of course there can be cases where the defect is reproducible but the
value from \textit{lastChange} is not. As in conventional breakpoint
debugging, we can only determine this empirically, by observing
multiple executions with different values. Unlike conventional
breakpoint debugging, implementations of \textit{lastChange} readily
aid in detecting such cases: as we work back through a series of
\textit{lastChange} requests we can compare values with previous
executions. Different values on different executions will signal that
the execution is not deterministic. In future we hope to compare
queries from successive executions as a tool for learning about
non-deterministic executions.

@@

We have presented the \textit{lastChange} algorithm and describe our
prototype implementation. Now we want to convince you that the
algorithm has practical value. Ultimately, practical value can only be
proven by developers in the field using the technique. This paper is
one step to convince implementers to put \textit{lastChange} in the
field.  Thus our goal here is a persuasive argument for potential implementors.

@@

We have shown how \textit{lastChange} identifies the prior point in
program execution where a program state value changes. Is it something
the developer needs to do?  Since ultimately programs are just
transformation of state values, debugging is ultimately backtracking
to find defects in program state change.  When a developer halts a
program on a breakpoint they have three kinds of information: a model
of the program in their head, a call stack showing what part of the
program is halted, and the state at that point (a combination of the
debugger's view of the program internal state and the output of the
program up to this point of execution). If the model does not match
the call stack or if no value in the viewable program state aligns
with the developers model for this point in execution, the developer
will seek another view by changing the breakpoint or the input
data. If a value appears incorrect, they may have a flash of insight
and know the defect. Otherwise they need to figure out what operation
causes the incorrect value. Here \textit{lastChange} takes over and addresses a key
part of the debugging process.

@@

We described our prototype implementation in Sec. 4. A fully usable
implementation would require access to the object id at the point of
object creation and \texttt{setPropertyChangeHook()}. Many
object-oriented runtimes provide object identifiers[] and some
provide access to object creation[]. In the particular case of
the Firefox Web browser we used for our prototype, the primary barrier
to practical implementation would likely be integration with the
increasingly sophisticated just in time compilers []. In the case of
tracking variable changes, \texttt{setVariableChangeHook()} can be implemented
similar to \texttt{setPropertyChangeHook()} if JavaScript scopes-like regular 
JavaScript objects-support \texttt{watch()} function.

@@

Among replay-based debuggers we compare to bdb \cite{Boothe} and
reverse watchpoint \cite{Maruyama}.  A bidirectional C debugger, bdb
employs a step counter to locate the requested point from the
beginning of execution. It relies on deterministic execution replay
(i.e., the same sequence of instructions in re-execution) and records
the results of non-deterministic system calls and re-injects them into
the program when it is replayed. It makes use of checkpoints to reduce
the time needed for re-execution.  Reverse watchpoint, is proposed by
Maruyama et al., analyses the execution and moves the debugger to the
last write access of a selected variable by re-executing the program
from the beginning\cite{Maruyama}.  Similar to bdb it relies on deterministic
replay and uses a counter to correctly locate a point in the next
execution. The main disadvantage of these approaches is requiring the
exactly the same executions. Even one instruction difference between
two executions leads to wrong results. On the other hand,
\textit{lastChange} doesn't require any special feature in the
re-execution and it is fit to  developers' everyday debugging
practice.

@@

Origin tracking of undefined and null values employing \textit{value piggybacking} technique proposed by
Bond et al. \cite{Bond}. This approach has two main limitations comparing to \textit{lastChange}.
First, it is limited to undefined and null values. Second, this approach dosen't return the last change
of a null variable but the first place that the null value is originated.

Two new directions in logging debuggers explore more detailed use of
the log and more effective logging approaches. WhyLine\cite{Ko}
provides visual interface to collected runtime information and let
developer to move on execution log using queries expressed in terms of
the programming objects. WhyLine stores the program user interface in
addition to program trace and provides answers to why and why not
questions to the user. Jive\cite{Czyz} depicts the history of
execution by a sequence diagram and lets user to query on events
database. Both tools suffer from the requirement of gathering tracing information before their unique capabilities can be used.
We imagine that the runtime queries we use in \textit{lastChange}  may be used to gather data incrementally for these kinds of debugging approaches.

>>>>

A bug is \textit{reproducible} for a developer when the developer can
start from a determined initial state, operate on the program with a
list of actions, and reproduce the symptoms of the bug. The details of
the execution can change each time we re-execute the buggy program,
but the buggy result is the same.  The reproducibility of the bug
means that the defect is very unlikely to depend on the order of
events during the execution. When the developer selects a value and
asks for \textit{lastChange}, they are expressing the hypothesis that
the value is related to the defect. Therefore the reproducibility of
the bug makes it very unlikely that the \textit{lastChange} will
depend upon order of events during execution.

Since \textit{lastChange} does not rely on determinism, some bugs that
confound breakpoint debuggers will be easier to find with
\textit{lastChange}.  Consider the example in 
Figure ~\ref{fig:counter-example}. It shows a function which processes an
array. The array contains numbers except one item which is
\texttt{undefined} and it causes a bug in line 30. There is a call to
\texttt{randomPermutation} function in line 11 which randomly
permutes the array item. So at every execution the
\texttt{undefined} item will be in a new place. Calling
\textit{lastChange} on \texttt{x} at the place the bug happens, gives a
point which shows line 14. Although this point exists at every
re-execution but it can not be identified by a conditional breakpoint.

@@

Of course there can be cases where the defect is reproducible but the
value from \textit{lastChange} is not. As in conventional breakpoint
debugging, we can only determine this empirically, by observing
multiple executions with different values. Unlike conventional
breakpoint debugging, implementations of \textit{lastChange} readily
help in detecting such cases: as we work back through a series of
\textit{lastChange} requests we can compare values with previous
executions. Different values on different executions will signal that
the execution is not deterministic. In future we hope to compare
queries from successive executions as a tool for learning about
non-deterministic executions.

@@

We have presented the \textit{lastChange} algorithm and described our
prototype implementation. Now we want to convince you that the
algorithm has practical value. Ultimately, practical value can only be
proven by developers in the field using the technique. This paper is
one step to convince implementers to put \textit{lastChange} in the
field.  Thus our goal here is a persuasive argument for potential implementors.

@@

We have shown how \textit{lastChange} identifies the prior point in
program execution where a program state value changes. Is it something
the developer needs to do?  Since ultimately programs are just
transformation of state values, debugging is ultimately backtracking
to find defects in program state change.  When a developer halts a
program on a breakpoint they have three kinds of information: a model
of the program in their mind, a call stack showing what part of the
program is halted, and the state at that point (a combination of the
debugger's view of the program internal state and the output of the
program up to this point of execution). If the model does not match
the call stack or if no value in the viewable program state aligns
with the developers model for this point in execution, the developer
will seek another view by changing the breakpoint or the input
data. If a value appears incorrect, they may have a flash of insight
and know the defect. Otherwise they need to figure out what operation
causes the incorrect value. Here \textit{lastChange} takes over and addresses a key
part of the debugging process.

@@

We have described our prototype implementation in Sec. 4. A fully usable
implementation would require access to the object id at the point of
object creation and \texttt{setPropertyChangeHook()}. Many
object-oriented runtimes provide object identifiers[] and some
provide access to object creation[]. In the particular case of
the Firefox Web browser we used for our prototype, the primary barrier
to practical implementation would likely be integration with the
increasingly sophisticated just in time compilers []. In the case of
tracking variable changes, \texttt{setVariableChangeHook()} can be implemented
similar to \texttt{setPropertyChangeHook()} if JavaScript scopes-like regular 
JavaScript objects-support \texttt{watch()} function.

@@

Among replay-based debuggers we compare to bdb \cite{Boothe} and
reverse watchpoint \cite{Maruyama}.  A bidirectional C debugger, bdb
employs a step counter to locate the requested point from the
beginning of execution. It relies on deterministic execution replay
(i.e., the same sequence of instructions in re-execution) and records
the results of non-deterministic system calls and re-injects them into
the program when it is replayed. It makes use of checkpoints to reduce
the time needed for re-execution.  Reverse watchpoint, is proposed by
Maruyama et al., analyses the execution and moves the debugger to the
last write access of a selected variable by re-executing the program
from the beginning\cite{Maruyama}.  Similar to bdb it relies on deterministic
replay and uses a counter to correctly locate a point in the next
execution. The main disadvantage of these approaches is that they require 
exactly the same executions. Even one instruction difference between
two executions leads to wrong results. On the other hand,
\textit{lastChange} doesn't require any special feature in the
re-execution and it is fit to  developers' everyday debugging
practice.

@@

Origin tracking of undefined and null values employing \textit{value piggybacking} technique proposed by
Bond et al. \cite{Bond}. This approach has two main limitations comparing to \textit{lastChange}.
First, it is limited to undefined and null values. Second, this approach dose not return the last change
of a null variable but the first place that the null value is originated.

Two new directions in logging debuggers explore more detailed use of
the log and more effective logging approaches. WhyLine\cite{Ko}
provides visual interface to collected runtime information and lets
developer moves on execution log using queries expressed in terms of
the programming objects. WhyLine stores the program user interface in
addition to program trace and provides answers to why and why not
questions to the user. Jive\cite{Czyz} depicts the history of
execution by a sequence diagram and lets user to query on events
database. Both tools suffer from the requirement of gathering tracing information before their unique capabilities can be used.
We imagine that the runtime queries we use in \textit{lastChange}  may be used to gather data incrementally for these kinds of debugging approaches.
